Open archaeology, open source?

### 2024-02-07 analysis/_06-network.qmd
Revise graph aesthetics

```{r fig-graph-repo-user}
#| fig-cap: Graph of open archaeology repositories and users connected by contributions. Darker edges indicate a great number of contributions. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
g_repo_user <- oarch_graph_repo_user |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(group = contrib, alpha = n_contrib)) +
  geom_point_interactive(aes(x, y, colour = cluster, shape = node_type, 
                             tooltip = node)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(shape = NULL) +
  theme(panel.background = element_rect(fill = "white"), 
        legend.position = "bottom")

@@

```{r fig-graph-repo-repo}
#| fig-cap: Graph of open archaeology repositories connected by common contributors. Darker edges indicate a great number of common contributors. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes repositories with only a single user.
g_repo_repo <- oarch_graph_repo_repo |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = cluster, tooltip = repo)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-user-user}
#| fig-cap: Graph of open archaeology users connected by contributions to common repositories. Darker edges indicate a great number of common repositories. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes users that have only contributed to a single repository.
g_user_user <- oarch_graph_user_user |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = cluster, tooltip = user)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

<!-- Is there a network? How big is it? -->
Our data shows that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository or user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-user and @fig-graph-repo-repo).

<!-- Clustering within the network -->
We delimited `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters that outline the topography of the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.

@@

```{r fig-repo-betweenness-age-length}
#| fig-cap: Repository centrality by age (left) and length (right). Centrality is measured by node betweenness weighted by number of contributions.
left <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  mutate(
    age = map_int(gh_commits, \(x) min(today() - date(x$datetime)))
  ) |>
  ggplot(aes(age, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "glm") +
  labs(x = "Repository age (days)", y = "Centrality") +
  theme_oarch()

right <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  mutate(
    length = map_int(gh_commits, nrow)
  ) |>
  ggplot(aes(length, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "glm") +
  labs(x = "Number of commits", y = NULL) +
  theme_oarch()

@@

<!-- Predictors of repo centrality -->
The repositories most central to the network as a whole (@tbl-repo-betweenness) include three lists and directories, including open-archaeo itself.
Three relate to making large data repositories accessible for analysis, and one is a very well-supported field recording application.
Community input is therefore centred on infrastructural projects, including those which index and publicize available tools and resources.
Moreover, three relate to radiocarbon data modelling and two relate to palaeoenvironment reconstruction, which reflects the fact that these have long been prominent foci of statistical software development in archaeology.

@@

```{r fig-repo-betweenness-category-platform}
#| fig-cap: Mean repository centrality by category (top) and platform (bottom). Centrality is measured by node betweenness weighted by number of contributions.
top <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  group_by(category) |>
  summarise(betweenness = mean(betweenness)) |>
  mutate(category = fct_reorder(category, betweenness)) |>
  ggplot(aes(category, betweenness)) +
  geom_col() +
  scale_y_log10() +
  coord_flip() +
  labs(y = NULL, x = NULL) +
  theme_oarch()

bottom <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  drop_na(platform) |>
  group_by(platform) |>
  summarise(betweenness = mean(betweenness)) |>
  arrange(-betweenness) |>
  slice(1:7) |>
  mutate(platform = fct_reorder(platform, betweenness)) |>
  ggplot(aes(platform, betweenness)) +
  geom_col() +
  scale_y_log10(labels = NULL) +
  coord_flip() +
  labs(y = "Centrality →", x = NULL) +
  theme_oarch()

@@

<!-- Predictors of user centrality -->
Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions (@fig-user-betweenness-contribs). 
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs [@baxter2012], and are increasingly precarious [@cornelius-bell2021].
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.

>>>>

```{r fig-graph-repo-user}
#| fig-cap: Graph of open archaeology repositories and users connected by contributions. Darker edges indicate a great number of contributions. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
g_repo_user <- oarch_graph_repo_user |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(group = contrib, alpha = n_contrib)) +
  geom_point_interactive(
    aes(x, y, colour = cluster, shape = node_type, tooltip = node),
    alpha = 0.8
  ) + 
  scale_edge_alpha(trans = "log", guide = guide_none(), range = c(0.05, 0.25)) +
  scale_colour_discrete(guide = guide_none()) +
  labs(shape = NULL) +
  theme(panel.background = element_rect(fill = "white"), 
        legend.position = "bottom")

@@

```{r fig-graph-repo-repo}
#| fig-cap: Graph of open archaeology repositories connected by common contributors. Darker edges indicate a great number of common contributors. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
g_repo_repo <- oarch_graph_repo_repo |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(
    aes(x, y, colour = cluster, tooltip = repo),
    alpha = 0.8
  ) + 
  scale_edge_alpha(guide = guide_none(), range = c(0.05, 0.25)) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-user-user}
#| fig-cap: Graph of open archaeology users connected by contributions to common repositories. Darker edges indicate a great number of common repositories. Node colour indicates membership of the largest clusters according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
g_user_user <- oarch_graph_user_user |>
  group_by(cluster) |>
  mutate(cluster_n = n()) |>
  ungroup() |>
  mutate(
    cluster_rank = dense_rank(desc(cluster_n)),
    cluster = if_else(cluster_rank <= 7, cluster, NA)
  ) |>
  ggraph(layout = "fr", niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(
    aes(x, y, colour = cluster, tooltip = user),
    alpha = 0.8
  ) + 
  scale_edge_alpha(guide = guide_none(), range = c(0.05, 0.25)) +
  scale_colour_discrete(guide = guide_none()) +
  theme(panel.background = element_rect(fill = "white"))

@@

Our data shows that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository or user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-user and @fig-graph-repo-repo).

We delimited `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters that outline the topography of the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.

@@

```{r fig-repo-betweenness-age-length}
#| fig-cap: Repository centrality by age (left) and length (right). Centrality is measured by node betweenness weighted by number of contributions.
left <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  mutate(
    age = map_int(gh_commits, \(x) min(today() - date(x$datetime)))
  ) |>
  ggplot(aes(age, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10(labels = NULL) +
  geom_smooth(method = "glm") +
  labs(x = "Days", y = "Log centrality →") +
  theme_oarch()

right <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  mutate(
    length = map_int(gh_commits, nrow)
  ) |>
  ggplot(aes(length, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10(labels = NULL) +
  geom_smooth(method = "glm") +
  labs(x = "Commits", y = NULL) +
  theme_oarch()

@@

The repositories most central to the network as a whole (@tbl-repo-betweenness) include three lists and directories, including open-archaeo itself.
Three relate to making large data repositories accessible for analysis, and one is a very well-supported field recording application.
Community input is therefore centred on infrastructural projects, including those which index and publicize available tools and resources.
Moreover, three relate to radiocarbon data modelling and two relate to palaeoenvironment reconstruction, which reflects the fact that these have long been prominent foci of statistical software development in archaeology.

@@

```{r fig-repo-betweenness-category-platform}
#| fig-cap: Mean repository centrality by category (top) and platform (bottom). Centrality is measured by node betweenness weighted by number of contributions.
top <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  group_by(category) |>
  summarise(betweenness = mean(betweenness)) |>
  mutate(category = fct_reorder(category, betweenness)) |>
  ggplot(aes(category, betweenness)) +
  geom_col() +
  scale_y_log10(labels = NULL) +
  coord_flip() +
  labs(y = NULL, x = NULL) +
  theme_oarch()

bottom <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  drop_na(platform) |>
  group_by(platform) |>
  summarise(betweenness = mean(betweenness)) |>
  arrange(-betweenness) |>
  slice(1:7) |>
  mutate(platform = fct_reorder(platform, betweenness)) |>
  ggplot(aes(platform, betweenness)) +
  geom_col() +
  scale_y_log10(labels = NULL) +
  coord_flip() +
  labs(y = "Log centrality →", x = NULL) +
  theme_oarch()

@@

Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions (@fig-user-betweenness-contribs). 
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs [@baxter2012], and are increasingly precarious [@cornelius-bell2021].
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.
